Thermal reactionomes reveal adaptive responses to thermal

### 2014-10-03 doc/ApTranscriptome_manuscript.Rmd

## Species differences in the number of thermally-responsive genes

The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at a 5% False Discovery Rate [@benjamini1995]. Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. In effect, `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive. This result is the same if we focus on only the transcripts that have a BLAST hit. However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] indicating that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive genes by expression type for *A. carolinensis* and *A. picea*. **Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other."}

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally-responsive genes by expression type for *A. carolinensis* and *A. picea*. **Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other.")
```

The number of genes that were responsive to cold temperatures was substantially greater than of the number responsive to hot temperatures. We grouped the thermally-responsive genes for each species into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). In both species,  the majority of genes were in the **Low** category (Table 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Table 1).

Gene set enrichment analysis (GSEA) of the responsive genes indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures. Specifically, out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` terms enriched (*P* < 0.01) for **Low** genes. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For *High* genes, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched terms, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process".  For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched and grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process". For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched and grouped into 3 clusters including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".

# Shifts in the expression patterns of thermally-responsive genes between species

Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species. While some of these genes were responsive in only one species (`r colSums(tt2)[which(colnames(tt2) == "NotResp")]` in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` in *A. picea* but not *A. carolinensis*), most of these were genes that changed from one expression category to another between the species (i.e. the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1).

In contrast to our prediction that more genes would be thermally-responsive in the warmer-climate species *A. carolinensis*, the total number of thermally-responsive genes did not differ between the species (...r t.test(Ap.type.table[1,1:4], Ap.type.table[2,1:4])...), but differed significantly in how they were distributed among the expression categories (Stuart-Maxwell test of marginal homogeneity ... ). Specifically, compared to a hypothetical homogenous distribution, there were an excess of **Intermediate** genes in *A. carolinensis* which disproportionatly switched to **Low** or **Bimodal** expression categories in *A. picea* (Fig. 1).

The molecular processes in this set of genes that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ... were related to metabolism. In addition, there were GO terms related to repair (e.g. "histone modification"), gene regulation (e.g. "positive regulation of cellular process") and cellular organization (e.g. "transport", "organelle organization"). These results indicate that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, shuts down metabolism while the colder-climate species, *A. picea*, increases metabolism.

```{r expression_change, echo=FALSE, warning = FALSE, fig.cap = "Matched expression type data for *A. carolinensis* (rows) and *A. picea* (columns). The color of each cell represent the deviation of the observed from the expected number of genes based on hypothetical equivalence of the marginal frequencies. Blue cells are those with significant excesses, while red cells are those with significant absences."}
#tt2

@@

## Species differences in the sensitivity of thermally-responsive genes to changes in temperature

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the mean temperatures of gene activation, defined as the temperatures at which the gene showed the greatest change in expression, for genes turned on at high or low temperatures, including *Bimodal* genes. Genes turned on at high temperatures in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~df=`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at low temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~df=`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

Similarly, for the **Intermediate** genes we compared the mean temperatures of downregulation of gene expression among species at both high (> 21°) and low (< 21°C) temperatures. Consistent with the results for up-regulated genes, we found that with increasing temperatures, the **Intermediate** genes were down-regulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~df=`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of down-regulation (*A. carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A. picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~df=`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`). Overall, these results indicate greater intertia of gene expression to changes in temperature for *A. carolinensis* than *A. picea*.

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*."}

T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = Species, group = Species, fill = Species)) +
    facet_grid(. ~ type) +
    geom_histogram(position = "dodge", binwidth = 3) +
    geom_density(alpha = 0.2, adjust = 3) + 
    scale_y_continuous(name="Density") +
    scale_x_continuous(name=expression("Temperature of gene activation"))
print(T_on_plot)
```

## Adaptive changes in basal expression are not linked to genes involved in short-term physiological acclimation

We looked for evidence of genetic assimilation, the process by which physiologically-relevant genes change from being environmentally-induced to basally-expressed, in multiple comparisons. First, we predicted that gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A. carolinensis* or high temperatures in *A. picea* should be higher in the opposite species (e.g. greater basal expression but lower inducibility). Instead, we found that gene expression at optimal temperatures (about 19.5°C) was always greater in *A. carolinensis* than in *A. picea*. This pattern was true for genes that had **Low** expression in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) **High** expression in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and **Intermediate** genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).

Second, for the genes that had **High** or **Bimodal** expression in *A. picea* but were not thermally-responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1), we predicted that mean expression level of the non-responsive *A. carolinensis* genes would be greater than the expression level near the optimum temperature (19.5°C) of the responsive *A. picea* genes. Again, we found no support for genetic assimilation as the non-responsive genes had lower mean expression levels.

@@

We examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Although `r resp.stress.responsive.count` genes were annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

@@

By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. We found that about 70% of the thermally-responsive genes (7,202) have non-linear responses to temperature.

@@

The patterns of gene expression between species differed substantially, with more genes down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Table 1, Fig. 1). The results of the gene set enrichment analysis give an indication of what might drive this difference. The genes that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly species, shuts-down metabolism to a greater extent than *A. picea* and so must rely on other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*.

@@

```{r max_exp_PDF, eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive genes in each species."}
# reshape data
Ap.df$Species <- ifelse(Ap.df$colony == "ApVT", "A. picea", "A. carolinensis")

@@

Given the differences in the number of thermally-responsive genes between the two species (Fig. 1), it is worth noting the similarity in expression patterns between species. In both species, there are 3-4 $\times$ more genes upregulated at cold than high temperatures  (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. At least among *Drosophila* taxa, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. In this light, for *A. picea* and *A. carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to cold temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

In contrast to the predictions of the genetic assimilation hypothesis that adaptation occurs via changes from plastically-induced to constitutive expression (e.g. genetic assimilation), we found lower basal gene expression in *A. picea* for genes that were up-regulated in response to cold temperatures (**Low** genes) in *A. carolinensis*. We also observed greater basal expression in *A. carolinensis* for **High** genes in *A. picea* as well as for genes that were **Intermediate** in both species. A potential explanation for the consistently higher basal gene expression in the warmer-environment species, *A. carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.

@@

The Trinity *de novo* transcriptome assembly for both species assembled together included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes. The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. We removed 5,675 genes identified as known contaminants, leaving 99,861 clean transcripts. Of these, 47% have best hits to genes from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).

To determine the putative function of the transcripts, we used functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies genes by Gene Ontology (GO) term assignment, enzyme identification and domain identification.

We used subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.

## Thermally-responsive genes

We quantified expression of each gene using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression. We included the contaminant genes identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other genes, but removed these from further analyses. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).

To identify genes that have significant changes in expression across the thermal gradient, we fit the regression model

>>>>

## Transcripts differ in thermal responsiveness between species

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally-responsive at a 5% False Discovery Rate [@benjamini1995]. Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. The result is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally-responsive in one species but not the other."}

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally-responsive in one species but not the other.")
```

Transcripts varied both qualitatively and quantitatively in their pattern of response to the temperature gradient. We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).

We found that the number of transcripts that responded to low temperatures is substantially greater than of that to high temperatures (Table 1). In both species, the majority of transcripts were in the **Low** category (Table 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate**.

Gene set enrichment analysis (GSEA) of the responsive transcripts indicated that the molecular responses to thermal extremes involve large-scale cellular and organizational changes. Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` terms enriched (*P* < 0.01) for **Low** transcripts. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For *High* transcripts, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched terms, with 8 of the 9 terms being involved in regulation of biosynthetic or metabolic process".  For **Bimodal** transcripts, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched and grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "transcript expression", "signaling", and "protein modification process". For **Intermediate** transcripts, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched and grouped into 3 clusters including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization".

# Shifts in the expression patterns of thermally-responsive transcripts between species

Of the thermally-responsive transcripts, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species. This was not due to a species difference in the overall number of responsive genes (...stat...). Instead, most of these were transcripts that changed from one expression category to another (i.e. the number of transcripts in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1). These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homotranscriptity ...)

**Intermediate** transcripts in *A. carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A. picea* than expected by chance (...stat...), while there was a significant deficit of the reverse shift of *A. picea* **Intermediate** transcripts to any upregulated category in *A. carolinensis* (Fig. 1). Similarly, both **High** and **Bimodal** transcripts in *A. picea* were more likely to be **Not Responsive** in *A. carolinensis* than expected, while the reverse was not significant (Fig. 1).

The molecular processes in the set of transcripts that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ... were related to metabolism. In addition, there were GO terms related to repair (e.g. "histone modification"), transcript regulation (e.g. "positive regulation of cellular process") and cellular organization (e.g. "transport", "organelle organization"). Altogether, these results suggest that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism while the colder-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

...GO terms **High** and **Bimodal** transcripts in *A. picea* that were **Not Responsive** in *A. carolinensis* ...

```{r expression_change, echo=FALSE, warning = FALSE, fig.cap = "Matched expression type data for *A. carolinensis* (rows) and *A. picea* (columns). The color of each cell represent the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies. Blue cells are those with significant excesses, while red cells are those with significant absences."}
#tt2

@@

## Species differences in the sensitivity of thermally-responsive transcripts to changes in temperature

To examine how the sensitivity of transcript expression to changes in temperature differed between the two species, we compared the mean temperatures of transcript activation, defined as the temperature at which the transcript showed the greatest positive change in expression. **High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes wre included for analysis at both high and low temperature activation. Transcripts turned on at high temperatures in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For low temperatures, there was no difference in the mean temperature of transcript activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of down-regulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures. Overall, the thresholds for down-regulation were shifted toward cooler temperatures than the corresponding thresholds for gene activation, with a shift of approximately ... at the higher temperature extremes and ...  at low temperature extremes. Consistent with the results for up-regulated transcripts, we found greater inertia of gene expression to changes in temperature for *A. carolinensis* than *A. picea* drive by a differential response to heat stress. With increasing temperatures, **Intermediate** transcripts were down-regulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~df=`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of down-regulation (*A. carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A. picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~df=`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).

**ADD PANEL TO FIG2 WITH DOWNREGULATION OF INTERMEDIATE TRANSCRIPTS**

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*."}

T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = Species, group = Species, fill = Species)) +
    facet_grid(. ~ type) +
    geom_histogram(position = "dodge", binwidth = 3) +
    geom_density(alpha = 0.2, adjust = 3) + 
    scale_y_continuous(name="Density") +
    scale_x_continuous(name=expression("Temperature of transcript activation"))
print(T_on_plot)
```

## Adaptive changes in basal expression are not linked to transcripts involved in short-term physiological acclimation

We tested the hypothesis of adaptation by genetic assimilation, the process by which transcripts change from being environmentally induced to constitutively expressed, in multiple comparisons. First, we predicted that consistent exposure to a temperature stressor should drive a shift from inducibility under stress conditions to continuously elevated expression (e.g. "frontloading", [@barshis2013]). Thus, the warm-adaptated *A. carolinensis* should display higher basal expression of **High** transcripts than *A. picea*, while the reverse should be true for **Low** transcripts.  Instead, we found that transcript expression at optimal temperatures (about 19.5°C) was always greater in *A. carolinensis* than in *A. picea*. This pattern was true for transcripts that had **High** expression in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`), **Low** expression in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`), and **Intermediate** transcripts in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).

In general, thermally-responsive transcripts in *A. carolinensis* had higher constitutive expression than those in *A. picea*.

The only exception was the set of genes that had **High** or **Bimodal** expression in *A. picea* but were not thermally-responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A. carolinensis*. Overall, we found no support for genetic assimilation as the primary mechanism for adaptive shifts in thermal tolerance.

@@

We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

@@

By using a regression-approach, we are able to quantitatively describe changes in transcript expression, rather than simply report the numbers of differentially-expressed transcripts. We found that about 70% of the thermally-responsive transcripts (7,202) have non-linear responses to temperature.

@@

The patterns of transcript expression between species differed substantially, with more transcripts down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Table 1, Fig. 1). The results of gene set enrichment analysis give an indication of what might drive this difference. The transcripts that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly species, shuts-down metabolism to a greater extent than *A. picea* and so must rely on other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*.

@@

```{r max_exp_PDF, eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive transcripts in each species."}
# reshape data
Ap.df$Species <- ifelse(Ap.df$colony == "ApVT", "A. picea", "A. carolinensis")

@@

Given the differences in the number of thermally-responsive transcripts between the two species (Fig. 1), it is worth noting the similarity in expression patterns between species. In both species, there are 3-4 $\times$ more transcripts upregulated at cold than high temperatures  (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. At least among *Drosophila* taxa, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. In this light, for *A. picea* and *A. carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to cold temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

In contrast to the predictions of the genetic assimilation hypothesis that adaptation occurs via changes from plastically-induced to constitutive expression (e.g. genetic assimilation), we found lower basal transcript expression in *A. picea* for transcripts that were up-regulated in response to cold temperatures (**Low** transcripts) in *A. carolinensis*. We also observed greater basal expression in *A. carolinensis* for **High** transcripts in *A. picea* as well as for transcripts that were **Intermediate** in both species. A potential explanation for the consistently higher basal transcript expression in the warmer-environment species, *A. carolinensis*, is that this species has a higher metabolic rate and thus requires greater transcript expression.

@@

The Trinity *de novo* transcriptome assembly for both species assembled together included 126,172 transcripts with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 transcripts. The total length was 63 million bp with an N~50~ length of 895 bp and a mean transcript size of 593 bp. Of the 105,536 filtered transcripts, 55,432 had hits to the NCBI-nr database. Of these, 38,711 transcripts mapped to GO terms, 1,659 transcripts were identified to an enzyme and 18,935 transcripts mapped to a domain with >50% coverage. We removed 5,675 transcripts identified as known contaminants, leaving 99,861 clean transcripts. Of these, 47% have best hits to transcripts from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).

To determine the putative function of the transcripts, we used functional annotation of the transcriptome assembly using the web-based tool `FastAnnotator` [@chen2012] which annotates and classifies transcripts by Gene Ontology (GO) term assignment, enzyme identification and domain identification.

## Thermally-responsive transcripts

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).

To identify transcripts that have significant changes in expression across the thermal gradient, we fit the regression model
